---
title: "Des chaines de Markov cachées couplées pour estimer la colonisation et la survie dans une métapopulation de plantes annuelles avec dormance"
author: "P.-O. Cheptou, S. Cordeau, S. Le Coz, N. Peyrard"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(doParallel)
library(optimr)
```

# 1. Load and transform data 

## 1.1 Load observations for the 7 study species

```{r}
set.seed(1)
#Setting the number of states 
#K observable states
K=5
#I hidden states
I=5
```

Getting the raw data for the 7 species of the study (density per square meter). Data are from Cordeau, S., Adeux, G., Meunier, D., Strbik, F., Dugué, F., Busset, H., Vieren, E., Louviot, G., & Munier-Jolain, N. (2020). Weed density of 7 major weeds in the long-term integrated weed management cropping system experiment of Dijon-Epoisses (2000-2017), https://doi.org/10.15454/M5P3LM. Portail Data INRAE.

```{r}
# save(data,file="data_epoisse.Rdata")
load("dat/data_epoisse.Rdata")
# raw data : data[[species]][patch,year]
```


```{r}
dtstate2=list()
oo=1
classS=list()
for(o in 1:7){
  dtstate2[[oo]]=array(rep(0,17*90),dim=c(90,17))
  classS[[oo]]=rep(0,K-2)
  oo=oo+1
}
names(classS)<-names(data)
names(dtstate2)<-names(data)
```

## 1.2 Define the bounds of the abundance classes of standing flora and transform raw data into abundance classes, for each species 

```{r}
oo=1
size_of_class=rep(0,length(names(dtstate2)))

for( o in names(dtstate2)){
  difference= max(log(data[[o]][which(data[[o]]!=0)]+1))
  interval=difference/4
  size_of_class[oo]=interval
  for(i in 1:90){
    for(j in 1:17){
      
      
      if(data[[o]][i,j]==0){dtstate2[[oo]][i,j]=1
      }else if(log(data[[o]][i,j]+1)<interval){dtstate2[[oo]][i,j]=2}
      else if(log(data[[o]][i,j]+1)<interval*2){dtstate2[[oo]][i,j]=3}
      else if(log(data[[o]][i,j]+1)<interval*3){dtstate2[[oo]][i,j]=4}
      else if(log(data[[o]][i,j]+1)>=interval*3){dtstate2[[oo]][i,j]=5}
      
    }
  }
  oo=oo+1
}
names(size_of_class)<-names(dtstate2)

size_of_class=size_of_class[c("ALOMY","CHEAL","SOLNI",'POLCO','AETCY','GALAP','POLAV')]

size_class_no_log=matrix(rep(0,5*7),nrow=7)
rownames(size_class_no_log)<-c("ALOMY","CHEAL","SOLNI",'POLCO','AETCY','GALAP','POLAV')
for(q in 1:7){
  
  size_class_no_log[q,4]=exp(size_of_class[q]*3-1)
  size_class_no_log[q,3]=exp(size_of_class[q]*2-1)
  size_class_no_log[q,2]=exp(size_of_class[q]-1)
}

size_class_no_log[,5]=rep(10000,7)

# dtstate2 gives the states (abundance class of standing flora) for each species in each patch for each year. 
# dtstate2[[species]][patch,year]
```


```{r}
### Load the data if you didn't run the transformation into abundance classes
# save(dtstate2,file='classes_uniforme_PIC_log(+1).Rdata')
load('dat/classes_uniforme_PIC_log(+1).Rdata')
```


## 1.3 Load the distances between the patches, and the crop types

```{r}
#loading cultd
#cultd gives the crop type for each patch each year between winter and summer.
#1=winter 2=summer 
load('dat/culture_epoisse_binaire.Rdata')

#loading diszon
load('dat/distance_entre_les_patchs.Rdata')

#missing data in patchs 64 and 74
index_patch_with_missing_info=c(64,71)
```


# 2. Compute spatial correlation between patchs (to build the neighbourhood)

You can directly go at the end of step 2 and load the results from Rdata files.

```{r}
diszone2=diszone[-index_patch_with_missing_info,-index_patch_with_missing_info]

correlation_results=list()

for(speci in c("ALOMY","CHEAL","SOLNI",'POLCO','AETCY','GALAP','POLAV')){
  


keepcor=array(rep(0,88*88),dim=c(88,88))
corecol=array(rep(0,88*16*88),dim=c(16,88,88))
nbr=0


data1=data[[speci]][-index_patch_with_missing_info,]

for(k in 1:88){
  corecol[,,k]=t(data1[,2:17])
  corecol[,k,k]=t(data1[k,1:16])
  
  keepcor[k,]=cor(corecol[,,k])[k,]
  
}
for(i in 1:88){
  for(j in 1:88){
    
   
    
    
    if(is.na(keepcor[i,j])){keepcor[i,j]=0}
    
    
    
  }}

for(i in 1:88){
  for(j in 1:88){
    if(i>j){diszone2[i,j]=-1}
  }}

ordr=sort(diszone2)
l=1

corki=list()
k=1

while(l<length(ordr)){
  if(l==1||ordr[l]!=ordr[l-1]){
   
    corki[[k]]=c(0,0)
    leschamps=which(ordr[l]==diszone2,arr.ind=TRUE)
    
    dd=dim(which(ordr[l]==diszone2,arr.ind=TRUE))[1]
   
    cumm=0
    for(oo in 1:dd ){
     
      cumm= keepcor[leschamps[oo,1],leschamps[oo,2]] +cumm + keepcor[leschamps[oo,2],leschamps[oo,1]]
      
    }

    corki[[k]]=c(cumm/(2*dd),ordr[l])
    names(corki[[k]])<-c('mean correlation','between patches distance')
    k=k+1
    
    
  }else{}
  
  l=l+1
  
}
correlation_results[[speci]]<-corki
}
```

Don't pay attention to the warnings. The warnings appear because we have less and less data when distance increases so we have a standard deviation of 0 sometimes. 

```{r}
suitcor<-list()
suitdist<-list()
plotlist<-list()
#fast plotting of the correlation results
for(o in c("ALOMY","CHEAL","SOLNI",'POLCO','AETCY','GALAP','POLAV') ){ 
  for(i in 1: 50){
    suitcor[i]<- correlation_results[[o]][[i]][1]
    suitdist[i]<- correlation_results[[o]][[i]][2]
  }

  plot(suitdist[2:30],suitcor[2:30],type='l',xlab='distance',ylab=c('correlation'),main=c(o))
}
```

Load the data if you didn't ran the correlations computation

```{r}
#save(correlation_results,file="correlation_results.Rdata")
load(file="dat/correlation_results.Rdata")
```


# 3. Estimation of MHMMDF model

Running time is long, 2-4 days ! You can directly go at the end of step 3 and load the results from Rdata files.

```{r}
#code used for the estimation of the MHMMDF model
source('code/EM-MHMM-DF-moy-inter.R')
```

```{r eval=FALSE}
#running estimations in parallel
nbr=3
#3 cores
registerDoParallel(nbr)

diszone3=diszone[-index_patch_with_missing_info,-index_patch_with_missing_info]

cultd3=cultd[-index_patch_with_missing_info,]

dtstate9=list()
#missing data in patchs  index_patch_with_missing_info
for( o in names(dtstate2)){
  dtstate9[[o]]=dtstate2[[o]][-index_patch_with_missing_info,]
}


#only 88 patchs as we deleted 2 due to missing data
CC=88
etatA9=list()
for(o in c("ALOMY","CHEAL","SOLNI",'POLCO','AETCY','GALAP','POLAV')){
  etatA9[[o]]=array(rep(0,17*88),dim=c(88,17))
  for(c in 1:CC){
    
    etatA9[[o]][c,]=field_to_vect_dist(c,dtstate9[[o]],K,diszone3)}}

#etatA9 gives the neighbors states 

#two estimations:  one with culture effect, the other one without
p = 100 #number max of EM iterations
for(variable_para_cult in 1:2)
{
ACSP3=list()
if(variable_para_cult==1){ culture_des_champs=matrix(rep(1, dim(cultd3)[1]* dim(cultd3)[2]), dim(cultd3))}else{ culture_des_champs=cultd3}

for(o in c("ALOMY","CHEAL","SOLNI",'POLCO','AETCY','GALAP','POLAV')){
  difsim=foreach(i=1:8) %dopar% EM_cult_dist_zi_neg(dtstate9[[o]][,1:16],etatA9[[o]][,1:16],CC,K,I,p,culture_des_champs[,1:16],culture_des_champs[,17],diszone3)
  print(o)
  #EM_cult_dist_zi_neg is the function used taking a long time
  vrai=-Inf
 
  for( j in 1:8){
    
    if(as.vector(difsim[[j]]$Vrai[2]) > vrai)
    {
      ACSP3[[o]]=difsim[[j]]
      vrai=difsim[[j]]$Vrai[2]
    }
  }
}
if(variable_para_cult==1){
  ACSP_nocult=ACSP3
}else{ ACSP_cult=ACSP3 }}
```

Save the results of model parameters estimates (nu, mu and tau)
```{r}
#save(ACSP_cult,file="EstimateurMHMMDF_cult_25_02_2020.Rdata")
#save(ACSP_nocult,file="EstimateurMHMMDF_25_02_2020.Rdata")
```

```{r eval=FALSE}
ACSP=ACSP_cult
ACSP=ACSP_nocult
```

Load the data if you didn't ran the parameters estimation

```{r}
load(file='dat/EstimateurMHMMDF_cult_25_02_2020.Rdata')
load(file='dat/EstimateurMHMMDF_25_02_2020.Rdata')
```

ACSP_cult contains the estimates of the MHMM-DF model with crop seasonality. ACSP_cult is used to compute (with step 4) the values in the table 'Tableau des probabilités de  survie  et de  sortie  de  dormance  pour  le modèle MHMM-DF qui tient compte de la saison de la culture locale'. 

ACSP_nocult contains the estimates of the MHMM-DF model without crop seasonality. ACSP_nocult is used to compute (with step 4) the values in  the table 'Tableau des probabilités de survie, de colonisation et de sortie de dormance pour le modèle non spatialisé et pour le MHMM-DF'

# 4. Calculate germination, seed survival, colonisation

Run time slow 1 day ! You can directly go at the end of step 4 and load the results from Rdata files.

```{r}
Estimateur_cult=ACSP_cult
Estimateur=ACSP_nocult
```

```{r eval = FALSE}
N = 100 # number of time steps in one simulation

set.seed(1)
#simulating the species trajectories across 17 years 100 times to get frequencies 
#used to calculate the ecological parameter from the model parameters 
#(germination, seed survival, colonisation) 
#running time is long because 7 species with 3 sets of parameters (winter, summer, none)
#for 17 years done 100 times 
for(w in 1:2){
    if(w== 1){
        ACSP = Estimateur_cult}
    else{
        ACSP = Estimateur}
    
    nombrecol=dim(ACSP[[1]]$paramestim$estimalpha)[2]
    
    lengspecie = length(c("ALOMY","CHEAL","SOLNI",'POLCO','AETCY','GALAP','POLAV'))
    
    freqX=array(rep(0,5*lengspecie*nombrecol),dim=c(lengspecie,5,nombrecol))
    freqY=array(rep(0,5*lengspecie*nombrecol),dim=c(lengspecie,5,nombrecol))
    rownames(freqX)<-c("ALOMY","CHEAL","SOLNI",'POLCO','AETCY','GALAP','POLAV')
    colnames(freqX)<-c('extinction','etat 2','etat 3','etat 4','etat 5')
    
    freqXX=array(rep(0,5*lengspecie*nombrecol),dim=c(lengspecie,5,nombrecol))
    freqY=array(rep(0,5*lengspecie*nombrecol),dim=c(lengspecie,5,nombrecol))
    rownames(freqXX)<-c("ALOMY","CHEAL","SOLNI",'POLCO','AETCY','GALAP','POLAV')
    colnames(freqXX)<-c('extinction','etat 2','etat 3','etat 4','etat 5')
    
    
    rownames(freqY)<-c("ALOMY","CHEAL","SOLNI",'POLCO','AETCY','GALAP','POLAV')
    colnames(freqY)<-c('extinction','etat 2','etat 3','etat 4','etat 5')
    
    for(hivete in 1 : nombrecol){
        print(hivete)
        for(z in 1:100){
            if(z%%10==0){print(z)}
            for(e in 1:lengspecie){
                
                C=88
                phi=ZIphi(ACSP[[e]]$paramestim$estimbeta[,hivete],I,K)
                A2=funcA_moy(ACSP[[e]]$paramestim$estimalpha[,hivete],I,K,C)
                pi=funcpi(ACSP[[e]]$paramestim$estimgamma,I)
                dist=diszone3
                
                for(q in 1:20){
                    
                    X= array(rep(0,C*(N+1)),dim=c(C,N+1))
                    Y= array(rep(0,C*(N+1)),dim=c(C,N))
                    #simulation with A2
                    t=runif(1)
                    X[,1]=length(which(cumsum(pi)<t))+1
                    
                    for(i in 1:N){
                        for(c in 1:C){
                            l= runif(1)
                            
                            if(l<phi[X[c,i],1]){ Y[c,i]=1
                            } else if(l <phi[X[c,i],1]+ phi[X[c,i],2]){ Y[c,i]=2
                            } else if(l <phi[X[c,i],1]+ phi[X[c,i],2]+ phi[X[c,i],3]){ Y[c,i]=3
                            } else if(l <phi[X[c,i],1]+ phi[X[c,i],2]+phi[X[c,i],3]+phi[X[c,i],4]){ Y[c,i]=4
                            } else { Y[c,i]=5}  
                        }
                        for(c in 1:C){
                            f=field_to_vect_dist(c,Y[,i],K,dist)
                            R=  runif(1)
                            
                            
                            if(R<A2[X[c,i],1,f]){ X[c,i+1]=1
                            } else if(R <A2[X[c,i],1,f]+ A2[X[c,i],2,f]){ X[c,i+1]=2
                            } else if(R <A2[X[c,i],1,f]+ A2[X[c,i],2,f]+ A2[X[c,i],3,f]){ X[c,i+1]=3
                            } else if(R <A2[X[c,i],1,f]+ A2[X[c,i],2,f]+A2[X[c,i],3,f]+A2[X[c,i],4,f]){ X[c,i+1]=4
                            } else{ X[c,i+1]=5}  
                            
                        }
                        
                        
                    }
                    for(i in 1:N){
                        y0=which(Y[,i]==1)
                        for(c in y0){
                            
                            
                            if(X[c,i]==1){
                                freqY[e,1,hivete]=freqY[e,1,hivete]+length(which(field_to_vect_dist(c,Y[,i],K,dist)%%K==1))
                                freqY[e,2,hivete]=freqY[e,2,hivete]+length(which(field_to_vect_dist(c,Y[,i],K,dist)%%K==2))
                                freqY[e,3,hivete]=freqY[e,3,hivete]+length(which(field_to_vect_dist(c,Y[,i],K,dist)%%K==3))
                                freqY[e,4,hivete]=freqY[e,4,hivete]+length(which(field_to_vect_dist(c,Y[,i],K,dist)%%K==4))
                                freqY[e,5,hivete]=freqY[e,5,hivete]+length(which(field_to_vect_dist(c,Y[,i],K,dist)%%K==0))}
                            
                            if(field_to_vect_dist(c,Y[,i],K,dist)%%K==1){
                                freqXX[e,1,hivete]=freqXX[e,1,hivete]+length(which(X[c,i]==1))
                                freqXX[e,2,hivete]=freqXX[e,2,hivete]+length(which(X[c,i]==2))
                                freqXX[e,3,hivete]=freqXX[e,3,hivete]+length(which(X[c,i]==3))
                                freqXX[e,4,hivete]=freqXX[e,4,hivete]+length(which(X[c,i]==4))
                                freqXX[e,5,hivete]=freqXX[e,5,hivete]+length(which(X[c,i]==5))}
                        }
                        
                    }
                    freqX[e,1,hivete]=freqX[e,1,hivete]+length(which(X==1))
                    freqX[e,2,hivete]=freqX[e,2,hivete]+length(which(X==2))
                    freqX[e,3,hivete]=freqX[e,3,hivete]+length(which(X==3))
                    freqX[e,4,hivete]=freqX[e,4,hivete]+length(which(X==4))
                    freqX[e,5,hivete]=freqX[e,5,hivete]+length(which(X==5))
                    
                    
                    
                    
                }
            }}}
    
    zziii<-list()
    AAAAAAQ<-list()
    AAAAAAQQ<-list()
    for(po in 1 : nombrecol){
        zziii[[po]]=(freqX[,,po]/rowSums(freqX[,-1,po]))[,-1]
        AAAAAAQ[[po]]= (freqX[,,po]/rowSums(freqX[,-1,po]))[,-1]
        AAAAAAQQ[[po]]= (freqX[,,po]/rowSums(freqX[,-1,po]))
        
        
    }
    survi=array(rep(0,lengspecie*nombrecol),dim=c(lengspecie,nombrecol)) 
    
    colon_moy=array(rep(0,lengspecie*nombrecol),dim=c(lengspecie,nombrecol)) 
    colon=array(rep(0,lengspecie*nombrecol),dim=c(lengspecie,nombrecol)) 
    floreleve=array(rep(0,lengspecie*nombrecol),dim=c(lengspecie,nombrecol)) 
    rownames(colon)<-c("ALOMY","CHEAL","SOLNI",'POLCO','AETCY','GALAP','POLAV')
    rownames(colon_moy)<-c("ALOMY","CHEAL","SOLNI",'POLCO','AETCY','GALAP','POLAV')
    rownames(floreleve)<-c("ALOMY","CHEAL","SOLNI",'POLCO','AETCY','GALAP','POLAV')
    probY=array(rep(0,lengspecie*5*nombrecol),dim=c(lengspecie,5,nombrecol))
    rownames(probY)<-c("ALOMY","CHEAL","SOLNI",'POLCO','AETCY','GALAP','POLAV')
    eAAAAAAQQ=array(rep(0,lengspecie*4*nombrecol),dim=c(lengspecie,4,nombrecol))
    
    
    rownames(survi)<-c("ALOMY","CHEAL","SOLNI",'POLCO','AETCY','GALAP','POLAV')
    if(nombrecol==2 ){ 
        colnames(survi)<-c('Winter','Summer')
        colnames(colon)<-c('Winter','Summer')
        colnames(colon_moy)<-c('Winter','Summer')
        colnames(floreleve)<-c('Winter','Summer')
    }
    else{}
```
    
Once the simulations are done we can calculate the seed survival, the colonisation and the germination
    
```{r eval = FALSE}
    for(hivete in 1 : nombrecol){
        
        for(e in 1 : lengspecie){
            
            
            C=88
            zziii[[hivete]][e,]=ZIphi(ACSP[[e]]$paramestim$estimbeta[,hivete],I,K)[-1,(-2:-5)]
            AAAAAAQ[[hivete]][e,]= funcA_moy(ACSP[[e]]$paramestim$estimalpha[,hivete],I,K,C)[-1,(-2:-5),1]
            survi[e,hivete]=(1-((diag(AAAAAAQ[[hivete]]%*%t((freqXX[,,hivete]/rowSums(freqXX[,-1,hivete]))[,-1]))))/(funcA_moy(ACSP[[e]]$paramestim$estimalpha[,hivete],I,K,C)[1,1,1]))[e]
            
            AAAAAAQQ[[hivete]][e,]= funcA_moy(ACSP[[e]]$paramestim$estimalpha[,hivete],I,K,C)[1,1,1:5]
            colon_moy[e,hivete]=(1-(diag(AAAAAAQQ[[hivete]]%*%t((freqY[,,hivete]/rowSums(freqY[,,hivete]))))))[e]
            
            floreleve[e,hivete]= (1-  (diag(zziii[[hivete]]%*%t(freqX[,-1,hivete]/rowSums(freqX[,-1,hivete])))))[e]
            
            probY[e,,hivete]=(freqX[,,hivete]/rowSums(freqX[,,hivete]))[e,]%*%ZIphi(ACSP[[e]]$paramestim$estimbeta[,hivete],I,K)
            
            eAAAAAAQQ[e,,hivete]= funcA_moy(ACSP[[e]]$paramestim$estimalpha[,hivete],I,K,C)[1,(-2:-5),2:5]
            colon[e,hivete]=(1-(diag(eAAAAAAQQ[,,hivete]%*%t((freqY[,,hivete]/rowSums(freqY[,-1,hivete]))[,-1]/(funcA_moy(ACSP[[e]]$paramestim$estimalpha[,hivete],I,K,C)[1,1,1])))))[e]
            
        }}
    
    if(nombrecol==2 ){ 
        results_culture<-list()
        results_culture$seed_survival<-survi
        results_culture$colonisation_voisin<-colon
        results_culture$colonisation_moy<-colon_moy
        results_culture$germination<-floreleve
        results_culture$probY<-probY
        results_culture$eAAAAAAQQ<-eAAAAAAQQ
        results_culture$AAAAAAQ<-AAAAAAQ
        results_culture$zziii<-zziii
        results_culture$AAAAAAQQ<-AAAAAAQQ
        results_culture$freqX<-freqX
        results_culture$freqXX<-freqXX
        results_culture$freqY<-freqY
        #save(results_culture,file="EstimateurMHMMDF_ecoparams_cult_25_02_2020.Rdata")
        load(file="dat/EstimateurMHMMDF_ecoparams_cult_25_02_2020.Rdata")
    }else{
        results_no_cult<-list()
        results_no_cult$seed_survival<-survi
        results_no_cult$colonisation_voisin<-colon
        results_no_cult$colonisation_moy<-colon_moy
        results_no_cult$germination<-floreleve
        results_no_cult$probY<-probY
        results_no_cult$eAAAAAAQQ<-eAAAAAAQQ
        results_no_cult$AAAAAAQ<-AAAAAAQ
        results_no_cult$zziii<-zziii
        results_no_cult$AAAAAAQQ<-AAAAAAQQ
        results_no_cult$freqX<-freqX
        results_no_cult$freqXX<-freqXX
        results_no_cult$freqY<-freqY
        #save(results_no_cult,file="EstimateurMHMMDF_ecoparams_25_02_2020.Rdata")
        load(file="dat/EstimateurMHMMDF_ecoparams_25_02_2020.Rdata")
    }
```

 
Load the data if you didn't ran the parameters estimation
```{r}
load(file="dat/EstimateurMHMMDF_ecoparams_25_02_2020.Rdata")
load(file="dat/EstimateurMHMMDF_ecoparams_cult_25_02_2020.Rdata")
```

results_culture gives the germination, seed survival, and colonisation for the MHMM-DF model with crop seasonality. results_culture is used in the table 'Tableau des probabilités de survie et de sortie de dormance pour le modèle MHMM-DF qui tient compte de la saison de la culture locale'. 

results_no_cult gives the germination, seed survival, and colonisation for the MHMM-DF model without crop seasonality. results_no_cult is used in the table 'Tableau des probabilités de survie, de colonisation et de sortie de dormance pour le modèle non spatialisé et pour le MHMM-DF'.


# 5. Calculating germination, seed survival and colonisation with Pluntz's model

Running time is fast 1 min ! You can directly go at the end of step 5 and load the results from Rdata files.

```{r}
## Load Mathieu Pluntz's estimation code
source('code/Markov chain parameters functions.R')
source('code/EM functions.R')
```

```{r}
set.seed(1)


done=list()

d=1
for(sp in names(dtstate2)){
  done[[sp]]=list()
  k=1
  for(cc in 1 : 88){
    if(cc%%9!=0){
      temp= dtstate2[[d]][cc,]
      temp[which(dtstate2[[d]][cc,]>1)]=1
      temp[which(dtstate2[[d]][cc,]==1)]=0
      done[[sp]][[k]]=temp
      k=k+1
    }
  }
  d=d+1
}

estBI90=list()

for(sp in c("ALOMY","CHEAL","SOLNI",'POLCO','AETCY','GALAP','POLAV')){
  estBI90[[sp]]=list()
  estBI90[[sp]]=EMestimation(done[[sp]],nIterations = 250)
}

mtrix3=array(rep(0,7*3),dim=c(7,3))
colnames(mtrix3)<-c('g','c','s')
rownames(mtrix3)<-c("ALOMY","CHEAL","SOLNI",'POLCO','AETCY','GALAP','POLAV')
k=1

for(sp in c("ALOMY","CHEAL","SOLNI",'POLCO','AETCY','GALAP','POLAV')){
  if(sp=="PLALA"){}else{
    mtrix3[k,1]= estBI90[[sp]]$param$g
    mtrix3[k,2]= estBI90[[sp]]$param$c
    mtrix3[k,3]= estBI90[[sp]]$param$s
   }
  k=k+1
}
```

Load the data if you didn't ran the parameters estimation

```{r}
#save(mtrix3,file="EstimateurPluntz_ecoparams_25_02_2020.Rdata")
load(file="dat/EstimateurPluntz_ecoparams_25_02_2020.Rdata")
mtrix3
```

mtrix3  gives the germination, seed survival and colonisation for species using Pluntz's model. mtrix3 is used in table 'Tableau des probabilités de survie, de colonisation et de sortie de dormance pour le modèle non spatialisé et pour le MHMM-DF'.

