---
title: "La régression linéaire généralisée sur composantes supervisées et ses extensions"
author: "F. Mortier, J. Chauvet, C. Trottier, G. Cornu, X. Bry"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---

# Initialization

## Load libraries

Be sure to use latest versions.
```{r}
#remotes::install_github("SCnext/SCGLR",force = TRUE)
library(SCGLR)

library(reshape2)

library(ggplot2)
library(maps)

library(future)
library(furrr)
library(progressr)
```

## Miscellaneous formatting

```{r}
# unicode values for common greek letters
greeks <- list(alpha='\u03b1', tau='\u03c4', sigma='\u03c3',
  beta='\u03b2', gamma='\u03b3', lambda='\u03bb', ell='\u2113')

# plot themes (ggplot)
theme_update(plot.title=element_text(hjust=0.5))
plot_theme <- theme(legend.title = element_text(size=12.5, face="bold"),
  legend.text = element_text(size=10, face="bold"), 
  axis.title.x = element_text(size=15, face="bold"),
  axis.title.y = element_text(size=15, face="bold"),
  axis.text = element_text(size=12))
map_theme <- plot_theme + theme(panel.background =element_rect(fill="black"))

# get congo basin country boundaries
congobasin <- map_data("world", region=c("Central African Republic", "Cameroon", 
  "Republic of Congo", "Gabon",
  "Democratic Republic of the Congo"))

# SCGLR plot styling
options(plot.SCGLR = list(
  title = "", # No title
  threshold = 0.8, # minimum correlation for being displayed (covariates & predictors)
  covariates.alpha = 0.7, # covariates are slightly transparent
  predictors = TRUE, # display also predictors (in red by default)
  predictors.labels.size = 5
))
```

## Explanatory and additional explanatory variables

```{r}
load("dat/genus2.RData")
genus <- genus2
n <- names(genus)
ny <- n[grep("^gen",n)]
nx <- n[-grep("^gen",n)] 
nx <- nx[!nx %in% c("geology","surface","center_x","center_y","inventory")]
na <- c("geology")
fam <- rep("poisson",length(ny))
form <- multivariateFormula(Y=ny, X=nx, A=na, data=genus)
```

## Dataset mixedgenus (for mixedSCGLR)

```{r}
# Additional explanatory variables
varaddi <- model.matrix(~factor(genus[,"geology"]))[,-1] 
colnames(varaddi) <- c("geol2","geol3","geol4","geol5")

# Dataset mixedgenus
mixedgenus <- list(
  Y = as.matrix(genus[, ny]),
  X = as.matrix(scale(genus[, nx])),
  AX = as.matrix(varaddi), 
  invent = as.factor(genus[, "inventory"]), 
  offset = matrix(rep(genus$surface,length(ny)), ncol=length(ny), byrow=FALSE))

# designXi and log(offset)
designXi <- model.matrix(~factor(mixedgenus$invent)-1)
colnames(designXi) <- paste("rand", 1:ncol(designXi), sep="")
loffset <- log(mixedgenus$offset)
```


## Division of the data into 5 folds

```{r}
nfolds <- 5
set.seed(112358)
folds <- vector("list", nfolds)
random <- mixedgenus$invent

for(j in 1:length(levels(random))){
  index_interest <- which(random==levels(random)[j])
  permutation <- sample(index_interest, length(index_interest), replace=F)
  divperm <- cut(seq(1,length(permutation)), breaks=nfolds, labels=FALSE)
  for(k in 1:nfolds){
    folds[[k]] <- c(folds[[k]], as.vector(permutation[divperm==k]))
  }
}

folds.scglr <- rep(NA,nrow(genus))
for(i in 1:nfolds){
  folds.scglr[folds[[i]]] <- i
}
```

## SCGLR

## 5-folds Cross-Validation for SCGLR

```{r}
s <- c(0.15,0.25,0.5)
l <- c(1,2,4)
design <- expand.grid(s,l)
design <- design[order(design$Var1),]
colnames(design) <- c("s","l")
rownames(design) <- paste("design", 1:nrow(design), ": ", sep=" ")
K <- 10

# start parallel processing
plan(multisession)

rmse_scglr <- with_progress({
  p <- progressor(nrow(design))
  furrr::future_pmap(design, function(s,l) {
    res <- try(scglrCrossVal(formula=form, 
                             data=genus2, 
                             family=fam, 
                             K=K,
                             folds=folds.scglr,
                             offset=genus2$surface,
                             method=methodSR(l=l, s=s, epsilon=1e-6),
                             crit=list(maxit=100)),
               silent = TRUE)
    p()
    res
  })
})

# stop parallel processing
plan(sequential)
```

## Results of the cross-validation

```{r}
rmse_scglr_geom <- 
  do.call(cbind, lapply(rmse_scglr, function(x) colMeans(log(x))))
rownames(rmse_scglr_geom) <- paste("H=", 0:10, sep="")
colnames(rmse_scglr_geom) <- 
  apply(design,1, 
    function(x) paste0(greeks$lambda, "=" ,x[1], ", ",
      greeks$ell, "=", x[2]))
data.recap <- as.data.frame(rmse_scglr_geom)
data.recap$id <- 0:10
plot_data.recap <- melt(data.recap, id.var="id")

plot.cv.scglr <- 
  ggplot(plot_data.recap, aes(x=id, y=value, group=variable, colour=variable)) +
  plot_theme+
  geom_point() + geom_line(size=1) +
  labs(x="Nombre de composantes (H)", y="Moyenne géométrique des RMSE",
    color = paste0("Paramètres de \ncompromis (", greeks$lambda, ")",
      "\net de localité \ndes faisceaux (", greeks$ell, ")")
  ) +
  geom_point(aes(x=7,y=1.106440), colour="red", size=4) +
  scale_x_continuous(breaks=seq(0,10,1)) + 
  scale_y_continuous(breaks=seq(1.1,1.4,0.05))
plot.cv.scglr
```

## Component planes SCGLR

```{r}
# Optimal parameters
tmp <- which(rmse_scglr_geom == min(rmse_scglr_geom), arr.ind=TRUE)
k_opt <- tmp[1]-1 
s_opt <- unlist(design[tmp[2],])[1]
l_opt <- unlist(design[tmp[2],])[2]

# SCGLR with optimal parameters
genus.scglr <- scglr(formula=form, data=genus, family=fam, offset=genus$surface,
  K=k_opt, method=methodSR(l=l_opt, s=s_opt, epsilon=1e-6),
  crit=list(maxit=100))

# Component planes
plot(genus.scglr, plane=c(1,2))
plot(genus.scglr, plane=c(1,3))
plot(genus.scglr, plane=c(2,3))

# SCGLR with parameters k_opt=7, s_opt=0.15, l_opt=4
genus.scglr2 <- scglr(formula=form, data=genus, family=fam, offset=genus$surface,
  K=7, method=methodSR(l=4, s=0.15, epsilon=1e-6),
  crit=list(maxit=100))

# Component planes
plot(genus.scglr2, plane=c(1,2))
plot(genus.scglr2, plane=c(1,3))
plot(genus.scglr2, plane=c(2,3))
```


## Map components

```{r}
# base map common to component maps
base_map <- ggplot(genus, aes(x=center_x, y=center_y)) +
  map_theme +
  labs(x="Longitude", y="Latitude") +
  geom_polygon(data=congobasin,aes(x=long, y=lat, group=group), fill="grey40", color="black") +
  # guides(fill=FALSE) + 
  coord_fixed(xlim=c(13,19), ylim=c(0.25,4.25))+
  scale_x_continuous(breaks=seq(13,19,1)) + 
  scale_y_continuous(breaks=seq(1,4,1)) +
  scale_fill_distiller(palette="Spectral") 

comp1_lonlat <- base_map +
  labs(fill="Comp 1") +
  geom_tile(aes(fill=genus.scglr$compr[,1]))
comp1_lonlat

comp2_lonlat <- base_map +
  labs(fill="Comp 2") +
  geom_tile(aes(fill=genus.scglr$compr[,2]))
comp2_lonlat

comp3_lonlat <- base_map +
  labs(fill="Comp 3") +
  geom_tile(aes(fill=genus.scglr$compr[,3]))
comp3_lonlat
```

## Theme SCGLR

## Definition of themes

```{r}
# THEME 1: Bio-physical variables
nx1 <- nx[-c(grep("^evi",nx), which(nx%in%c("MIR","NIR")))]

# THEME 2: Variables describing the photosynthetic activity
nx2 <- nx[c(grep("^evi",nx), which(nx%in%c("MIR","NIR")))] 
```

## Backward selection

```{r}
form_theme <- multivariateFormula(ny, nx1, nx2, A=na)
genus.thm <- scglrThemeBackward(formula=form_theme, 
                                data=genus, 
                                family=fam,
                                offset=genus$surface,
                                folds=folds.scglr,
                                H=c(6,6),
                                method=methodSR(l=l_opt, s=s_opt, epsilon=1e-6))
```

## Graph of the backward selection

```{r}
cv.thm <- as.matrix(genus.thm$cv_path[order(length(genus.thm$cv_path):1)])
colnames(cv.thm) <- c("theme-SCGLR")
row.names(cv.thm) <- c("0_0", "1_0","2_0","3_0","4_0","5_0", 
  "5_1","5_2","5_3","5_4","5_5","5_6", "6_6")
data.recap <- as.data.frame(cv.thm)
data.recap$id <- row.names(cv.thm)
plot_data.recap <- melt(data.recap,id.var="id")

plot.cv.themescglr <-
  ggplot(plot_data.recap, aes(x=id, y=value, group=variable, colour=variable)) +
  plot_theme+
  geom_point(color='steelblue') + geom_line(color='steelblue', size=1) +
  labs(x="Nombre de composantes par thème", y="Moyenne géométrique des RMSE", 
    color="Méthode") +
  geom_point(aes(x=8,y=1.121), colour="red", size=4) +
  scale_y_continuous(limits=c(1.1,1.42), breaks=seq(1.1,1.4,0.1))
plot.cv.themescglr
```


## Component planes THEME-SCGLR

```{r}
# THEME-SCGLR with optimal parameters
genus.thm <- scglrTheme(formula=form_theme, data=genus, family=fam, 
  offset=genus$surface,
  H=c(5,3),
  method=methodSR(l=l_opt, s=s_opt, epsilon=1e-6))

# Component planes (THEME 1)
plot(genus.thm$themes[[1]], plane=c(1,2))
plot(genus.thm$themes[[1]], plane=c(1,3))
plot(genus.thm$themes[[1]], plane=c(2,3))

# Component planes (THEME 2)
plot(genus.thm$themes[[2]], plane=c(1,2))
plot(genus.thm$themes[[2]], plane=c(1,3))
plot(genus.thm$themes[[2]], plane=c(2,3))
```


# Mixed SCGLR

## Parameter grid for the 5-folds cross-validation

```{r}
val_k <- 1:10 
val_s <- c(0.15) 
val_l <- c(4) 
for_k <- rep(val_k, each=length(val_s)*length(val_l))
for_s <- rep(  rep(val_s, each=length(val_l)), length(val_k)  )
for_l <- rep(val_l, length(val_s)*length(val_k))
par_ksl <- cbind(for_k, for_s, for_l)
```

## 5-folds CROSS-VALIDATION (parallel computing)

```{r}
# start parallel processing
plan(multisession)

error.CV <- lapply(seq(nfolds), function(i) {
  cat("Fold ", i, sep="")
  cat("\n")
  cal <- (1:nrow(mixedgenus$Y))[ -folds[[i]] ]
  val <- folds[[i]]
  
  Y_cal <- as.matrix(mixedgenus$Y[cal,])
  X_cal <- as.matrix(mixedgenus$X[cal,])
  AX_cal <- as.matrix(mixedgenus$AX[cal,])
  designXi_cal <- as.matrix(designXi[cal,])
  loffset_cal <- as.matrix(loffset[cal,])
  random_cal <- random[cal]
  
  Y_val <- mixedgenus$Y[val,]
  X_val <- mixedgenus$X[val,]
  AX_val <- mixedgenus$AX[val,]
  designXi_val <- designXi[val,]
  loffset_val <- loffset[val,]
  
  nbtriplet <- nrow(par_ksl)
  
  BigMatrice <- furrr::future_map(1:nbtriplet, function(jj) {
    tryCatch({
      tmp <- kCompRand(Y=Y_cal, X=X_cal, AX=AX_cal, 
        random=random_cal, loffset=loffset_cal,
        family=rep("poisson",ncol(mixedgenus$Y)),
        init.sigma=rep(1,ncol(mixedgenus$Y)), 
        init.comp="pca",
        k=as.numeric(par_ksl[jj,1]),
        method=methodSR("vpi",
          s=as.numeric(par_ksl[jj,2]),
          l=as.numeric(par_ksl[jj,3]),
          maxiter=1000, 
          epsilon=10^-6, bailout=1000))
      
      xnew <- cbind(1, X_val, AX_val, as.matrix(designXi_val))
      betanew <- as.matrix(rbind(as.matrix(tmp$beta),
        as.matrix(tmp$blup)))
      pred <- SCGLR:::multivariatePredictGlm(
        Xnew=xnew,
        family=rep("poisson",ncol(mixedgenus$Y)),
        beta=betanew,
        offset=exp(loffset_val)
      )
      
      tmperror <- colMeans((Y_val-pred)^2/pred)
      return(c(tmperror))
    }, error = function(e) {
      return(NULL) # drop triplet in case of error
    }) 
  })

  do.call(cbind, BigMatrice)
})

# stop parallel processing
plan(sequential)

# ATTENTION marche pas s'il y a eu une erreur car les dimensions ne seront plus les mêmes !
recap.error.CV <- Reduce("+", error.CV, init=0)/nfolds
colnames(recap.error.CV) <- paste("tripletPAR", 1:nrow(par_ksl), sep="")
moy.geom.err <- apply(recap.error.CV, 2, function(x) mean(log(x)))
```

## Results of the cross-validation (SCGLR vs MIXED-SCGLR)

```{r}
par_ksl.err <- cbind(par_ksl, moy.geom.err)
select.s <- val_s 
select.l <- val_l 
mat.select <- 
  par_ksl.err[(par_ksl.err[,2]%in%select.s) & (par_ksl.err[,3]%in%select.l),]
matrix.recap <- matrix(NA, length(select.s)*length(select.l), length(val_k))
for(k in val_k){
  matrix.recap[,k] <- mat.select[mat.select[,1]==k, 4]
}
matrix.recap <- rbind(matrix.recap, rmse_scglr_geom[2:11,3])
row.names(matrix.recap) <- c("mixedSCGLR", "SCGLR")
colnames(matrix.recap) <- paste("H=", val_k, sep="")

data.recap <- as.data.frame(t(matrix.recap))
data.recap$id <- val_k
plot_data.recap <- melt(data.recap, id.var="id")

plot.cv.mixedscglr <- 
  ggplot(plot_data.recap, aes(x=id, y=value, group=variable, colour=variable)) +
  plot_theme+
  geom_point() + geom_line(size=1) +
  labs(x="Nombre de composantes (H)", y="Moyenne géométrique des RMSE", 
    color="Méthode") +
  scale_x_continuous(breaks=seq(1,10, by=1))
plot.cv.mixedscglr
```

## Component planes MIXED-SCGLR

```{r}
optimal.triplet <- par_ksl.err[which.min(par_ksl.err[,4]), -4]
k.opt <- optimal.triplet[1]
s.opt <- optimal.triplet[2]
l.opt <- optimal.triplet[3]

withRandom.opt <- 
  kCompRand(Y=mixedgenus$Y, X=mixedgenus$X, AX=mixedgenus$AX,
    family=rep("poisson", ncol(mixedgenus$Y)),
    random=mixedgenus$invent, loffset=log(mixedgenus$offset),
    init.sigma=rep(1, ncol(mixedgenus$Y)), init.comp="pca",
    k=k.opt, method=SCGLR::methodSR("vpi", l=l.opt, s=s.opt,
      maxiter=1000, epsilon=10^-6, 
      bailout=1000))

plot(withRandom.opt, plane=c(1,2))
plot(withRandom.opt, plane=c(1,3))
plot(withRandom.opt, plane=c(2,3))

withRandom.opt2 <- 
  kCompRand(Y=mixedgenus$Y, X=mixedgenus$X, AX=mixedgenus$AX,
    family=rep("poisson", ncol(mixedgenus$Y)),
    random=mixedgenus$invent, loffset=log(mixedgenus$offset),
    init.sigma=rep(1, ncol(mixedgenus$Y)), init.comp="pca",
    k=3, method=SCGLR::methodSR("vpi", l=4, s=0.15,
      maxiter=1000, epsilon=10^-6, 
      bailout=1000))

plot(withRandom.opt2, plane=c(1,2))
plot(withRandom.opt2, plane=c(1,3))
plot(withRandom.opt2, plane=c(2,3))
```

# Spearman correlations 

## SCGLR

```{r}
pred.scglr <- matrix(0, nrow(genus), length(ny))
plots <- 1:nrow(genus)

for (i in 1:5) {
  print(paste(i, "/", 5))
  plots_val <- plots[folds.scglr == i]
  plots_cal <- plots[folds.scglr != i]
  genus.scglr2 <- scglr(form, data=genus, family=fam, offset=genus$surface,
    subset=plots_cal,
    K=7,
    method=methodSR(l=4, s=0.15, bailout=100,
      maxiter=100, epsilon=1e-6))
  
  ## Validation matrix
  x_new <- model.matrix(form, data=genus[plots_val,], rhs=1:length(form)[2])
  
  ## Predicting abundances on validation dataset
  pred.scglr[plots_val,] <- 
    multivariatePredictGlm(x_new, family=fam, beta=as.matrix(genus.scglr2$beta),
      offset=genus$surface[plots_val])
}
res.scglr <- diag(cor(pred.scglr, genus[,ny], method="spearman"))
```

## MIXED-SCGLR

```{r}
pred.mixedscglr <- matrix(0, nrow(genus), length(ny))
plots <- 1:nrow(genus)

for (i in 1:5) {
  print(paste(i, "/", 5))
  plots_val <- plots[folds[[i]]]
  plots_cal <- plots[-folds[[i]]]
  fit.mixedscglr <- 
    kCompRand(Y=mixedgenus$Y[plots_cal,], 
      family=rep("poisson", ncol(mixedgenus$Y)),
      X=mixedgenus$X[plots_cal,], AX=mixedgenus$AX[plots_cal,],
      random=random[plots_cal], loffset=loffset[plots_cal,],
      init.sigma=rep(1,ncol(mixedgenus$Y)), init.comp="pca",
      k=3,method=methodSR("vpi", s=0.15, l=4, maxiter=1000, 
        epsilon=10^-6, bailout=1000))
  ## Validation matrix
  xnew.mixed <- cbind(1, mixedgenus$X[plots_val,], mixedgenus$AX[plots_val,], 
    as.matrix(designXi[plots_val,]))
  ## Predicting abundances on validation dataset
  betanew.mixed <- as.matrix(rbind(as.matrix(fit.mixedscglr$beta),
    as.matrix(fit.mixedscglr$blup)))
  pred.mixedscglr[plots_val,] <- 
    SCGLR:::multivariatePredictGlm(Xnew=xnew.mixed ,
      family=rep("poisson",ncol(mixedgenus$Y)),
      beta=betanew.mixed, 
      offset=exp(loffset[plots_val,]))
}
res.mixedscglr <- diag(cor(pred.mixedscglr, genus[,ny], method="spearman"))

spearman <- round(cbind(res.scglr, res.mixedscglr), 2)
row.names(spearman) <- paste0("gen",seq_len(nrow(spearman)))
spearman
```

# Abundance maps

## Data

```{r}
data <- data.frame(y=genus$gen9,
                   predscglr=pred.scglr[,9],
                   predmixed=pred.mixedscglr[,9])
data$long <- genus$center_x
data$lat <- genus$center_y

# base map
lim <- 100
base_map <- ggplot(data,aes(x=long,y=lat)) +
  map_theme+
  labs(x="Longitude",y="Latitude")+
  geom_polygon(data=congobasin,aes(x=long, y=lat, group=group), fill="grey40", color="black") +
  coord_fixed(xlim=c(13,19), ylim=c(0.25,4.25))+
  scale_x_continuous(breaks=seq(13,19,1)) +
  scale_y_continuous(breaks=seq(1,4,1)) +
  scale_fill_gradientn(colours=terrain.colors(10),
                       values=(1-log(c(1:100))/max(log(c(1:100)))),
                       limits=c(0,lim))
```

## Observed abundance

```{r}
real.abundance <- 
  base_map + labs(fill="Vraie\nabondance") +
  geom_tile(aes(fill=y))
real.abundance
```

## Predictions - SCGLR

```{r}
pred.abundance.scglr <- 
  base_map + labs(fill="Prédictions\nSCGLR") +
  geom_tile(aes(fill=predscglr))
pred.abundance.scglr
```

## Predictions - mixedSCGLR

```{r}
pred.abundance.mixedscglr <- 
  base_map + labs(fill="Prédictions\nmixed-SCGLR") +
  geom_tile(aes(fill=predmixed))
pred.abundance.mixedscglr
```



